Unsupervised Learning for Identifying Events in Active Target Experiments

### 2020-08-05 doc/Manuscript/article.tex
Updates from Overleaf

In addition to the performance measures reported in table \ref{tab:clstr_vgg}, it is interesting to observe which samples are wrongly assigned. To investigate this problem, we compute the matrices $\mathbf{W}$ as shown in table \ref{tab:contingency}. From these tables, we can infer which classes are more or less entangled with others. The results for each data set is shown in figures \ref{fig:clster_confmat_sim} \ref{fig:clster_confmati_filt} and \ref{fig:clster_confmati_raw}. We observe that the proton class is consistently assigned in a pure cluster.. For example, consider the row corresponding to the proton class in figure \ref{fig:clster_confmati_filt}. The column corresponding to the largest entry in the proton row has zero other predicted classes in it.

This high-quality cluster also appears in the clustering of raw data. From figure \ref{fig:clster_confmati_raw}, we observe that there is a high purity proton cluster. In contrast to the filtered data we observe that the deterioration in performance can largely be ascribed to the algorithm creating a proton + other cluster and a carbon + other cluster.

@@

We repeat this analysis using a Principle Component Analysis (PCA) dimensionality reduction\footnote{
PCA is a common techinque to find the significant variations in data by projecting the data along a subset of its covariance matrix eigenvectors \cite{Marsland2009}
} on the latent space of the VGG16 model. This is done to estimate to what degree the class separating information is encoded in the entirety of the latent space, or in some select regions. The results from the PCA analysis, using the top 100 principal components , were virtually identical to the results sans the PCA. This result indicates that the class-encoding information is contained in a minority of the axes of variation in the data.
%and so we omit them for brevity.

To further investigate the clusters presented in the matrix in figures \ref{fig:clster_confmati_filt} and \ref{fig:clster_confmati_raw}, we sample examples from the proton samples belonging to different clusters for the filtered and full data.
We display these examples in figures \ref{fig:filtered_vgg_clster_repr} and \ref{fig:full_vgg_clster_repr} respectively. Form these figures we gleam that the more noisy proton events are being clustered together with the amorphous other class. This is in line with the intuitive expectation from viewing the clustering tables in figures \ref{fig:clster_confmati_filt} and \ref{fig:clster_confmati_raw}.

\begin{figure}
\includegraphics[width=\textwidth]{custom_work/Filteredexamples.pdf}
\caption[Filtered proton samples by cluster belonging]{Illustrating a sample of proton events from different $K$-means clusters from the filtered data set. Each row belongs to a single cluster corresponding to the filtered confusion matrix in figure \ref{fig:clster_confmati_filt}}\label{fig:filtered_vgg_clster_repr}
\end{figure}

\begin{figure}
\includegraphics[width=\textwidth]{custom_work/Rawexamples.pdf}
\caption[Full proton samples by cluster belonging]{Illustrating a sample of proton events from different $K$-means clusters from the raw data set. Each row belongs to a single cluster corresponding to the full confusion matrix in figure \ref{fig:clster_confmati_raw}}\label{fig:full_vgg_clster_repr}
\end{figure}

@@

In the previous section we demonstrated a powerful, but rather naive, clustering technique for AT-TPC track identification. To build on this result we will in this section explore the application of the mixture of autoencoders (MIXAE) algorithm introduced in section \ref{sec:mixae}.
For details on hyper-parameter tuning and the experimental procedure for training the MIXAE algorithm see appendix \ref{ax: mixae}

After hyperparameter tuning, each highest performing model is run $10$ times. The results are listed in table \ref{tab:clstr_mixae}.
We observe that, while the algorithm can achieve very strong performance, the performance varies. In some cases the MIXAE model converges to a seemingly good configuration, based on its unsupervised training goals. However, when inspecting its clustering performance against labelled data that seemingly good model does no better than random. 
This happens more frequently with raw data, indicating an interaction with the noise levels present in the events.

@@

As with the VGG16 + KMeans approach we wish to further investigate the clustering results. Taking the best performing MIXAE model on filtered and raw data we tabulate the clusters against their labels. These tables are present in figures \ref{fig:mixae_confmat_filtered} and \ref{fig:mixae_confmat_raw} for filtered and raw data, respectively.

@@

On filtered data the highest performing MiXAE model achieves strong separation of the other-class, but curiously creates two proton-majority clusters.

@@

The most striking result we present in this work is the success of the $K$-means approach. As noted by \cite{Aggarwal} distance measures become less informative in higher dimensional spaces, but the $K$-means algorithm clusters events well in a very high-dimensional space.
Another surprise is the stability of the $K$-means algorithm. A stability we attribute to the quality of the VGG16 latent space in creating class-separating sub-spaces. While the separations are not perfect, the stability and quality of the proton track identification create solid empirical grounding for applying this approach to other active target experiments.

It is also interesting to compare and contrast the clustering results from the MIXAE model with those of the VGG16$+$$K$-means. In particular, the discrepancy in stability is worth noting. While the top performing MIXAE runs outperform the $K$-means approach, its reliability suffers. However, the high performance achieved shines a light on valuable potential research into more tailored models for unsupervised track identification.

@@

In addition to the results presented in this section, we performed clustering with a number of different algorithms included in the \lstinline{scikit-learn} package. None of them provided any notable differences from the $K$-means results or were significantly worse. Notably, the DBSCAN algorithm \cite{Ester96adensity-based}\cite{Bergstra2012} failed to provide any useful clustering results. We find this important as one of the significant drawbacks of $K$-means, and the deep clustering algorithm presented in section \ref{sec:mixae}, is that they are all dependent on pre-determining the number of clusters. This is not the case for DBSCAN.

Additionally, we considered the deep convolutional embedded clustering (DCEC) by \citet{Guo2017} as well as the MIXAE method introduced by \citet{Zhang}. While we were able to reproduce the authors' results on their data, the DCEC algorithm proved unable to cluster AT-TPC data in our implementation. However, this provides valuable insight as the seeds of the clusters are constructed by a $K$-means algorithm. This insight contrasts with our positive results from applying a pre-trained classification model with $K$-means and highlights potentially significant differences in models trained on a supervised or unsupervised objective for clustering tasks in nuclear physics.

>>>>

In addition to the performance measures reported in table \ref{tab:clstr_vgg}, it is interesting to observe which samples are wrongly assigned. To investigate this problem, we compute the matrices $\mathbf{W}$ as shown in table \ref{tab:contingency}. From these tables, we can infer which classes are more or less entangled with others. The results for each data set is shown in figures \ref{fig:clster_confmat_sim} \ref{fig:clster_confmati_filt} and \ref{fig:clster_confmati_raw}. We observe that the proton class is consistently assigned in a pure cluster. For example, consider the row corresponding to the proton class in Fig.~\ref{fig:clster_confmati_filt}. The column corresponding to the largest entry in the proton row has zero other predicted classes in it.

This high-quality cluster also appears in the clustering of raw data. From Fig.~\ref{fig:clster_confmati_raw}, we observe that there is a high purity proton cluster. In contrast to the filtered data we observe that the deterioration in performance can largely be ascribed to the algorithm creating a proton plus another cluster and a carbon plus another cluster.

@@

We repeat this analysis using a Principle Component Analysis (PCA) dimensionality reduction\footnote{
PCA is a common techinque to find the significant variations in data by projecting the data along a subset of its covariance matrix eigenvectors \cite{vidal2016,Marsland2009}
} on the latent space of the VGG16 model. This is done to estimate to what degree the class separating information is encoded in the entirety of the latent space, or in some select regions. The results from the PCA analysis, using the top 100 principal components, were virtually identical to our previous results not containing  the PCA analysis. This in an interesting observation which indicates that the class-encoding information is contained in a minority of the axes of variation in the data.
%and so we omit them for brevity.

To further investigate the clusters presented in the matrix in figures \ref{fig:clster_confmati_filt} and \ref{fig:clster_confmati_raw}, we study examples from the proton samples belonging to different clusters for the filtered and full data.
We display these examples in Figs.~\ref{fig:filtered_vgg_clster_repr} and \ref{fig:full_vgg_clster_repr}, respectively. Form these figures we observe that the more noisy proton events are being clustered together with the amorphous other class events. This is in line with the intuitive expectation from viewing the clustering results in Figs.~\ref{fig:clster_confmati_filt} and \ref{fig:clster_confmati_raw}.

\begin{figure}
\includegraphics[width=\textwidth]{custom_work/Filteredexamples.pdf}
\caption[Filtered proton samples by cluster belonging]{Illustration of  a sample of proton events from different $K$-means clusters originating from the filtered data set. Each row belongs to a single cluster corresponding to the filtered confusion matrix in Fig.~\ref{fig:clster_confmati_filt}}\label{fig:filtered_vgg_clster_repr}
\end{figure}

\begin{figure}
\includegraphics[width=\textwidth]{custom_work/Rawexamples.pdf}
\caption[Full proton samples by cluster belonging]{Illustration of  a sample of proton events from different $K$-means clusters originating from the raw data set. Each row belongs to a single cluster corresponding to the full confusion matrix in Fig.~\ref{fig:clster_confmati_raw}}\label{fig:full_vgg_clster_repr}
\end{figure}

@@

In the previous section we demonstrated a powerful, but rather naive, clustering technique for AT-TPC track identification. To build on this result we will in this section explore the application of the mixture of autoencoders (MIXAE) algorithm introduced in section \ref{sec:mixae}.
For details on hyper-parameter tuning and the experimental procedure for training the MIXAE algorithm see the appendix \ref{ax: mixae}.

With the best set of hyperparameters, each highest performing model is thereafter run $10$ times. The results are listed in table \ref{tab:clstr_mixae}.
We observe that, while the algorithm can achieve a very strong performance, the performance varies. In some cases the MIXAE model converges to a seemingly good configuration, based on its unsupervised training goals. However, when inspecting its clustering performance against labelled data, the seemingly good model does no better than a model based on random selection. 
This happens more frequently with the raw data, indicating an interaction with the noise levels present in the events.

@@

As with the VGG16 + $K$-Means approach we wish to further investigate the clustering results. Taking the best performing MIXAE model on filtered and raw data we tabulate the clusters against their labels. These tables are present in figures \ref{fig:mixae_confmat_filtered} and \ref{fig:mixae_confmat_raw} for filtered and raw data, respectively.

@@

On filtered data the highest performing MIXAE model achieves strong separation of the other-class, but curiously creates two proton-majority clusters.

@@

The most striking result we present in this work is the success of the $K$-means approach. As noted by \cite{Aggarwal}, distance measures become less informative in higher dimensional spaces, but the $K$-means algorithm clusters events well in a very high-dimensional space.
Another surprise is the stability of the $K$-means algorithm. We attribute this stability to the quality of the VGG16 latent space in creating class-separating sub-spaces. While the separations are not perfect, the stability and quality of the proton track identification create solid empirical grounding for applying this approach to other active target experiments.

It is also interesting to compare and contrast the clustering results from the MIXAE model with those of the VGG16 with the  $K$-means approach. In particular, the discrepancy in stability is worth noting. While the top performing MIXAE runs outperform the $K$-means approach, its reliability suffers. However, the high performance achieved indicates that this may represent a valuable potential research path into more tailored models for unsupervised track identification.

@@

In addition to the results presented in this section, we performed clustering with a number of different algorithms included in the \lstinline{scikit-learn} package. None of them provided any notable differences from the $K$-means results or were significantly worse. Notably, the DBSCAN algorithm \cite{Ester96adensity-based}\cite{Bergstra2012} failed to provide any useful clustering results. We find this important as one of the significant drawbacks of $K$-means, and the deep clustering algorithm presented in section \ref{sec:mixae}, is that they depend  on pre-determining the number of clusters. This is not the case for DBSCAN.

Additionally, we considered the deep convolutional embedded clustering (DCEC) approach by \citet{Guo2017} as well as the MIXAE method introduced by \citet{Zhang}. While we were able to reproduce the authors' results on their data, the DCEC algorithm proved unable to cluster AT-TPC data in our implementation. However, this provides valuable insight as the seeds of the clusters are constructed by a $K$-means algorithm. This insight contrasts with our positive results from applying a pre-trained classification model with $K$-means and highlights potentially significant differences in models trained on a supervised or unsupervised objective for clustering tasks in nuclear physics.
